---
title: "Code along, Boston BlueBikes"
author: "Dr. Cannataro"
output: html_document
---

# Goal: Model bike rides vs. weather data 

> Which weather variable explains the most variation within bike rides?

# Input data... 

```{r load_in_packages}
library(tidyverse)
# make sure to install the tidymodels package
# remember, you only need to install packages once, 
# so no need to keep the function in the .Rmd document. 
# Here, I comment it out
# install.packages("tidymodels")

# but we still need to load it in every time with library()
library(tidymodels)
```

## Bike data 

```{r testing_a_file}
# read in one of the files. 
# Wow, read_csv can read in .zip files! 
jan_rides <- readr::read_csv(file = "data/bluebikes/201801-hubway-tripdata.zip")

# Look, the `birth year` variable is thought to be a character string. 
# This will bite us later. 
jan_rides$`birth year` %>% unique()

# See the NULL? Keep that in mind. 
```

### Iterate through the files



```{r all_my_files}

# we want full.names=T because we need the file names to have the directory
my_files <- dir(path = "data/bluebikes/",full.names = T)

```


#### Using a for loop

```{r iterate_through_my_files, message=F}

# using a for loop

# same thing
seq_along(my_files) 
1:length(my_files)


# create an empty data set
bike_data <- NULL


# for each number in the total number of files... 
# e.g., file 1, 2, 3, ... 
for(file_index in seq_along(my_files)){
  
  # read in this file
  this_df <- readr::read_csv(file = my_files[file_index])
  
  # bind it to our main data frame
  bike_data <- rbind(bike_data, this_df)
  
  # tell me where you are in the loop 
  # (this is useful because sometimes loops break / get stuck and you can tell where)
  # (it is also fun)
  message(paste("I just read and pasted the ",file_index," file"))
  
  
}


# check the top and the bottom of my data
head(bike_data)
tail(bike_data)

```


#### Using map

```{r using_map, message=F}

# remember how above we saw that NULL in the birth year caused 
# birth year to be assumed to be a  character vector for some months?
# let's correct that by providing this option to read_csv.
bike_data_2 <- purrr::map_dfr(my_files, read_csv, na = "NULL")

# check the top and the bottom... 
head(bike_data_2)
tail(bike_data_2)
# looks the same! 
```

Yay! 


## Count the trips per day
```{r}

# count the number of trips per day, and rename the variable
total_bike_rides <- bike_data %>% 
  mutate(date = lubridate::date(starttime)) %>%
  count(date) %>%
  rename(total_bike_rides = "n") 

```


```{r}
# take a look at the data
ggplot(total_bike_rides) + 
  geom_line(aes(x=date,y=total_bike_rides))
```

# Getting the weather data

```{r}
# load the weather data in.
# downloaded directly from the website
# https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USW00014739/detail
# and cleaned with the cleaning script in the data/weather directory. 
boston_weather <- readr::read_csv(file = "data/weather/weather_data_clean.csv")
```


# Joining it all together

```{r}

# join the columns by the date
bike_and_weather_data <- left_join(x = total_bike_rides,y = boston_weather, by = c("date" = "DATE"))

# save for future use/sharing with the class! 
readr::write_rds(x = bike_and_weather_data, file = "data/bike_and_weather_data.rds")

## This is how we load it back in if we are starting from here. 
# bike_and_weather_data <- readr::read_rds(file = "data/bike_and_weather_data.rds")

```


# Exploring some variables

```{r}
ggplot(bike_and_weather_data,aes(y=total_bike_rides, x=Snowfall)) + 
  geom_point() + 
  geom_smooth(method = "lm")


ggplot(bike_and_weather_data,aes(y=total_bike_rides, x=Precipitation)) + 
  geom_point() + 
  geom_smooth(method = "lm")

ggplot(bike_and_weather_data,aes(y=total_bike_rides, x=Average_temperature)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```


# Building a function to make our plots and give us R-squared


```{r}

# define a variable name here so we can step through the guts of our function to test it

variable_name <- "Precipitation"
dot_color <- "blue"

get_r_squared <- function(variable_name, dot_color, line_color="orange"){
  
  # now we use aes_string because we are providing our variable names as strings
  my_plot <- ggplot(bike_and_weather_data,
                    aes_string(y="total_bike_rides", x=variable_name)) + 
    geom_point(color = dot_color) + 
    geom_smooth(method = "lm", color=line_color)
  
  # create a temporary dataframe with our data so we can model
  data_to_fit <- tibble(explanatory = pull(bike_and_weather_data[,variable_name]), 
                        response = pull(bike_and_weather_data[,"total_bike_rides"]))
  
  # create our model
  my_model <- linear_reg() %>%
    set_engine("lm") %>%
    fit(response ~ explanatory, data = data_to_fit)
  
  # glance(my_model)$r.squared
  
  # return a list with our R-squared and our plot
  return(
    list(my_r_squared = glance(my_model)$r.squared,
         my_plot = my_plot)
  )
  
}


```

## testing the function

```{r}

function_output <- get_r_squared(variable_name = "Precipitation",dot_color = "blue",line_color = "red")

function_output$my_r_squared

function_output$my_plot

function_output <- get_r_squared(variable_name = "Average_temperature", dot_color="blue")

function_output$my_r_squared

function_output$my_plot

```

# Finding the variable with the highest R-squared

```{r}


# create a table to store our data
r_squared_table <- tibble(variable = colnames(bike_and_weather_data)[3:10], r_squared = NULL)

# using a loop to fill in our table
for(row_ind in 1:nrow(r_squared_table)){
  
  r_squared_table[row_ind,"r_squared"] <- get_r_squared(variable_name = pull(r_squared_table[row_ind,"variable"]),dot_color = "blue")$my_r_squared
  
}


# display our table 
r_squared_table %>%
  arrange(desc(r_squared))

```

