---
title: "Nobel winners"
author: ""
output: github_document
---

```{r load-packages, message=FALSE}
library(tidyverse)
```


- ECLearn > Assignments > `in-class exercise, Nobels and sales + Data import` > open `nobels-csv.Rmd` and knit.
- Read in the `nobels.csv` file from the `data-raw/` folder.
- Split into two (STEM and non-STEM): 
  - Create a new data frame, `nobel_stem`, that filters for the STEM fields 
(Physics, Medicine, Chemistry, and Economics).
  - Create another data frame, `nobel_nonstem`, that filters for the remaining 
fields.  
- Write out the two data frames to `nobel-stem.csv` and `nobel-nonstem.csv`, 
respectively, to `data/`.

**Hint:** Use the `%in%` operator when `filter()`ing.



Let's first load the data:

```{r label-me1, eval=FALSE}
nobel <- ___(___)
```

Then let's split the data into two:

```{r label-me2, eval=FALSE}
# stem laureates
___ <- nobel %>%
  filter(___)

# non-steam laureates
___ <- nobel %>%
  filter(___)
```

And finally write out the data:

```{r label-me3}
# add code for writing out the two data frames here
```
