---
title: "Modeling, part 2: Model checking and linear models with multiple predictors"
subtitle: "<br><br> IDDS 2201: Data Analytics, Module 13"
author: "Vincent L. Cannataro, Ph.D."
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---


```{r child = "../setup.Rmd"}
```


```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ggtext)
library(knitr)
library(patchwork)
library(kableExtra)
set.seed(1234)
options(dplyr.print_min = 10, dplyr.print_max = 6)
```



class: middle

# Model checking

---

## Data: Paris Paintings

```{r message=FALSE}
pp <- read_csv("data/paris-paintings.csv", na = c("n/a", "", "NA"))
```

- Number of observations: `r nrow(pp)`
- Number of variables: `r ncol(pp)`

---

## "Linear" models

- We're fitting a "linear" model, which assumes a linear relationship between our explanatory and response variables.
- But how do we assess this?

---

## Graphical diagnostic: residuals plot

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "residuals-plot", echo = FALSE, warning = FALSE, out.width = "60%"}
```
]
.panel[.panel-name[Code]
```{r residuals-plot, fig.show="hide"}
ht_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(Height_in ~ Width_in, data = pp)

ht_wt_fit_aug <- augment(ht_wt_fit$fit) #<<

ggplot(ht_wt_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted height", y = "Residuals")
```
]

.panel[.panel-name[Augment]
```{r}
ht_wt_fit_aug
```
]
]

---

## More on `augment()`

```{r}
glimpse(ht_wt_fit_aug)
```


---

## Looking for...

- Residuals distributed randomly around 0
- With no visible pattern along the x or y axes

```{r out.width = "60%", echo=FALSE}
df <- tibble(
  fake_resid = rnorm(1000, mean = 0, sd = 30),
  fake_predicted = runif(1000, min = 0, max = 200)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Fan shapes**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_resid = c(rnorm(100, mean = 0, sd = 1), 
                 rnorm(100, mean = 0, sd = 15), 
                 rnorm(100, mean = 0, sd = 25), 
                 rnorm(100, mean = 0, sd = 20), 
                 rnorm(100, mean = 0, sd = 25), 
                 rnorm(100, mean = 0, sd = 50), 
                 rnorm(100, mean = 0, sd = 35), 
                 rnorm(100, mean = 0, sd = 40),
                 rnorm(200, mean = 0, sd = 80)),
  fake_predicted = seq(0.2, 200, 0.2)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Groups of patterns**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(0.2, 200, 0.2),
  fake_resid = c(
    rnorm(500, mean = -20, sd = 10),
    rnorm(500, mean = 10, sd = 10)
  )
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Residuals correlated with predicted values**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(0.2, 200, 0.2),
  fake_resid = fake_predicted + rnorm(1000, mean = 0, sd = 50)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Any patterns!**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(-100, 100, 0.4),
  fake_resid = -5*fake_predicted^2 - 3*fake_predicted + 20000 + rnorm(501, mean = 0, sd = 10000)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

.question[
What patterns does the residuals plot reveal that should make us question whether a linear model is a good fit for modeling the relationship between height and width of paintings?
]

```{r out.width = "60%", echo=FALSE}
ggplot(ht_wt_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted height", y = "Residuals")
```

---

class: middle

# Exploring linearity

---

## Data: Paris Paintings

```{r echo=FALSE, out.width = "70%"}
ggplot(data = pp, aes(x = price)) +
  geom_histogram(binwidth = 1000) +
  labs(title = "Prices of paintings")
```

---

## Price vs. width

```{r echo=FALSE, out.width = "70%", warning = FALSE}
ggplot(data = pp, aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(x = "Width (in)", y = "Price (livres)")
```

---

## Focus on paintings with `Width_in < 100`

That is, paintings with width < 2.54 m

```{r}
pp_wt_lt_100 <- pp %>% 
  filter(Width_in < 100)
```

---

## Price vs. width

.question[
Which plot shows a more linear relationship?
]

.small[

.pull-left[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Price vs. width", 
    subtitle = "For width < 100 in",
    x = "Width (inches)", 
    y = "Price (livres)"
  )
```
]

.pull-right[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Log(price) vs. width", 
    subtitle = "For width < 100 in",
    x = "Width (inches)", 
    y = "Log(price) (log livres)"
  )
```
]

]

---

## Price vs. width, residuals

.question[
Which plot shows a residuals that are uncorrelated with predicted values from the model? Also, what is the unit of the residuals?
]

.pull-left[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
price_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(price ~ Width_in, data = pp_wt_lt_100)
price_wt_fit_aug <- augment(price_wt_fit$fit)

ggplot(data = price_wt_fit_aug, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Price vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted price (livres)", 
    y = "Residuals"
  )
```
]
.pull-right[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
pp_wt_lt_100 <- pp_wt_lt_100 %>% 
  mutate(log_price = log(price))

lprice_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Width_in, data = pp_wt_lt_100)
lprice_wt_fit_aug <- augment(lprice_wt_fit$fit)

ggplot(data = lprice_wt_fit_aug, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Log(Price) vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted log(price) (log livres)", 
    y = "Residuals"
  )
```
]

---

## Transforming the data

- We saw that `price` has a right-skewed distribution, and the relationship between price and width of painting is non-linear.

--
- In these situations a transformation applied to the response variable may be useful.

--
- In order to decide which transformation to use, we should examine the distribution of the response variable.

--
- The extremely right skewed distribution suggests that a log transformation may 
be useful.
- log = natural log, $ln$
- Default base of the `log` function in R is the natural log: <br>
`log(x, base = exp(1))`

---

## Logged price vs. width

.question[
How do we interpret the slope of this model?
]

```{r echo=FALSE, message=FALSE, out.width="60%"}
ggplot(data = pp_wt_lt_100, mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "#8E2C90", se = FALSE) +
  labs(x = "Width (in)", y = "Log(price) (log livres)")
```

---

## Models with log transformation

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(log(price) ~ Width_in, data = pp_wt_lt_100) %>%
  tidy()
```

---


## Interpreting the slope

$$ \widehat{log(price)} = 4.67 + 0.0192 Width $$

--
- For each additional inch the painting is wider, the log price of the painting is expected to be higher, on average, by 0.0192 livres.

--
- which is not a very useful statement...

---

## Working with logs

- Subtraction and logs: $log(a) − log(b) = log(a / b)$

--
- Natural logarithm: $e^{log(x)} = x$

--
- We can use these identities to "undo" the log transformation

---

## Interpreting the slope

The slope coefficient for the log transformed model is 0.0192, meaning the log price difference between paintings whose widths are one inch apart is predicted to be 0.0192 log livres.

--

.question[
Using this information, and properties of logs that we just reviewed, fill in the blanks in the following alternate interpretation of the slope:

>For each additional inch the painting is wider, the price of the painting is expected to be `___` , on average, by a factor of `___`.
]


---

>For each additional inch the painting is wider, the price of the painting is expected to be `___` , on average, by a factor of `___`.


$$ log(\text{price for width x+1}) - log(\text{price for width x}) = 0.0192 $$

--

$$ log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right) = 0.0192 $$

--

$$ e^{log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right)} = e^{0.0192} $$

--

$$ \frac{\text{price for width x+1}}{\text{price for width x}} \approx 1.02 $$

--

For each additional inch the painting is wider, the price of the painting is expected to be higher, on average, by a factor of 1.02.


---

## A note on log transforming exponential relationships

- The log transformation of exponential growth... 

--

$$ y = a \times 2^x $$

--

$$ log(y) = log(a \times 2^x)$$
--

$$ log(y) = log(a) + log(2^x)$$

--


$$ log(y) = log(a) + x \times log(2) $$
--

- ... results in a linear relationship 

---

## A note on log transforming exponential relationships

```{r, echo=F}

# data from https://github.com/nytimes/covid-19-data/
nytimes_covid_data <- readr::read_csv(file = "data/us-states.csv")

nytimes_covid_data %>% 
  group_by(date) %>%
  summarize(total_cases = sum(cases)) -> 
  total_covid_cases


```





```{r, echo=F}
ggplot(total_covid_cases %>% filter(date < "2020-04-01", date > "2020-02-05"),
       aes(x=date, y=total_cases)) + 
  geom_point(size=3) +
  geom_line() + 
  labs(y="Total COVID19 cases in the USA",x="Date") + 
  theme_minimal()
```



---

## A note on log transforming exponential relationships

```{r, echo=F}
library(scales)
total_covid_cases %>% 
  filter(date < "2020-04-01", 
         date > "2020-02-05") %>%
  mutate(date_since = as.numeric(date - min(date))) %>%
  ggplot(aes(x=date, y=total_cases)) + 
  geom_point(size=3) +
  geom_line() + 
  labs(y="Total COVID19 cases in the USA, log scale",x=" Date") + 
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) + 
  theme_minimal()
```



---

## Recap

- Non-constant variance is one of the most common model violations, however it is usually fixable by transforming the response (y) variable.

--
- The most common transformation when the response variable is right skewed is the log transform: $log(y)$, especially useful when the response variable is 
(extremely) right skewed.

--
- This transformation is also useful for variance stabilization.

--
- When using a log transformation on the response variable the interpretation of 
the slope changes: *"For each unit increase in x, y is expected on average to be higher/lower <br> by a factor of $e^{b_1}$."*

--
- Another useful transformation is the square root: $\sqrt{y}$, especially 
useful when the response variable is counts.

---

## Transform, or learn more?

- Data transformations may also be useful when the relationship is non-linear
- However in those cases a polynomial regression may be more appropriate
+ This is beyond the scope of this course, but you are welcomed to try it for your final project, and I would be happy to provide further guidance

---

## Aside: when $y = 0$

In some cases the value of the response variable might be 0, and

```{r}
log(0)
```

--

The trick is to add a very small number to the value of the response variable for these cases so that the `log` function can still be applied:

```{r}
log(0 + 0.00001)
```

---

class: middle

# The linear model with multiple predictors

---


## Data: Book weight and volume

The `allbacks` data frame gives measurements on the volume and weight of 15 books, some of which are paperback and some of which are hardback

.pull-left[
- Volume - cubic centimeters
- Area - square centimeters
- Weight - grams
]
.pull-right[
.small[
```{r echo=FALSE}
library(DAAG)
as_tibble(allbacks) %>%
  print(n = 15)
```
]
]

.footnote[
.small[
These books are from the bookshelf of J. H. Maindonald at Australian National University.
]
]

---

## Book weight vs. volume

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight)) +
  geom_point(alpha = 0.7, size = 3)
```
]

---

## Book weight vs. volume and cover

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight, color = cover, shape = cover)) +
  geom_point(alpha = 0.7, size = 3) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]

---

## Interpretation of estimates

```{r echo=FALSE}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```

--
- **Slope - volume:** *All else held constant*, for each additional cubic centimeter books are larger in volume, we would expect the weight to be higher, on average, by 0.718 grams.

--
- **Slope - cover:** *All else held constant*, paperback books are weigh, on average, by 184 grams less than hardcover books.

--
- **Intercept:** Hardcover books with 0 volume are expected to weigh 198 grams, on average. (Doesn't make sense in context.)


---

## Main vs. interaction effects

.question[
Suppose we want to predict weight of books from their volume and cover type 
(hardback vs. paperback). Do you think a model with main effects or 
interaction effects is more appropriate? Explain your reasoning.

**Hint:** Main effects would mean rate at which weight changes as volume 
increases would be the same for hardback and paperback books and interaction 
effects would mean the rate at which weight changes as volume 
increases would be different for hardback and paperback books.
]

---

```{r book-main-int, echo=FALSE, out.width="65%", fig.asp = 0.8}
book_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks)
book_main_fit_aug <- augment(book_main_fit$fit)

book_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover + volume*cover, data = allbacks)
book_int_fit_aug <- augment(book_int_fit$fit)

p_main <- ggplot() +
  geom_point(book_main_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_main_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Main effects, parallel slopes", 
       subtitle = "weight-hat = volume + cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_int <- ggplot() +
  geom_point(book_int_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_int_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Interaction effects, not parallel slopes", 
       subtitle = "weight-hat = volume + cover + volume * cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_main /
  p_int +
  plot_layout(guides = "collect")
```

---

## In pursuit of Occam's razor

- Occam's Razor states that among competing hypotheses that predict equally well, the one with the fewest assumptions should be selected.

--
- Model selection follows this principle.

--
- We only want to add another variable to the model if the addition of that variable brings something valuable in terms of predictive power to the model.

--
- In other words, we prefer the simplest best model, i.e. **parsimonious** model.

---

.pull-left[
```{r ref.label = "book-main-int", echo = FALSE, out.width="100%", fig.asp = 0.8}
```
]
.pull-right[
.question[
Visually, which of the two models is preferable under Occam's razor?
]
]

---

## R-squared

- $R^2$ is the percentage of variability in the response variable explained by 
the regression model.

```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```

--
- Clearly the model with interactions has a higher $R^2$.

--
- However using $R^2$ for model selection in models with multiple explanatory variables is not a good idea as $R^2$ increases when **any** variable is added to the model.

---

## Adjusted R-squared

... a (more) objective measure for model selection

- Adjusted $R^2$ doesn't increase if the new variable does not provide any new 
information or is completely unrelated, as it applies a penalty for number of 
variables included in the model.
- This makes adjusted $R^2$ a preferable metric for model selection in multiple
regression models.

---

## Comparing models

.pull-left[
```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```
]
.pull-right[
```{r}
glance(book_main_fit)$adj.r.squared
glance(book_int_fit)$adj.r.squared
```
]

--

.pull-left-wide[
.small[
- Is R-sq higher for int model?
```{r}
glance(book_int_fit)$r.squared > glance(book_main_fit)$r.squared 
```

- Is R-sq adj. higher for int model?

```{r}
glance(book_int_fit)$adj.r.squared > glance(book_main_fit)$adj.r.squared
```
]
]



---

# Attributions

Some of the material in this slide deck was inspired by and/or partially adapted from several open-source data science resources, including

- The [datasciencebox](https://github.com/rstudio-education/datascience-box) by Mine Çetinkaya-Rundel [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/) 
- The [R for Data Science](https://r4ds.had.co.nz/) textbook by Hadley Wickham & Garrett Grolemund [CC BY-NC-ND 3.0 US](https://creativecommons.org/licenses/by-nc-nd/3.0/us/) 
- Artwork by @allison_horst [CC BY-SA 4.1](https://creativecommons.org/licenses/by-sa/4.0/) 




